---
title: "Viral clearance"
author: "James Watson"
date: "20/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, cache.comments = FALSE,
                      echo = F, include = TRUE, 
                      fig.width = 8, fig.height = 8,
                      fig.pos = 'H',dev = 'png', dpi = 300)

library(rstan)
library(matrixStats)
library(mgcv)
library(survival)
```


This RMarkdown plots the output of the model fits and simulations

```{r load_data_model_fits}
# this loads dataset and runs stan models
source('functions.R')
source('run_models.R')
writeLines(sprintf('There are a total of %s infection episodes', length(unique(CT_data$PersonID))))

range(table(CT_data$PersonID))
median(table(CT_data$PersonID))

IDs = as.numeric(names(which(sort(table(CT_data$PersonID))>2)))

D_maxs = c(5,7,14)
D_min = 0
```


## Time to clearance

There are multiple ways to define time to clearance. We use time to first CT value equal to 40.

```{r time_to_clear}
CT_data$ID = CT_data$PersonID
CT_data$Trt_effect = CT_data$Vaccinated
CT_data$Day = round(CT_data$time)

CT_data_time_clear = 
  time_to_clear(log10_vl = CT_data$log10_vl,
                time = CT_data$time,
                Trt_effect = CT_data$Vaccinated, 
                ID = CT_data$ID,
                log10_cens_vl = unique(CT_data$log10_cens_vl))
hist(CT_data_time_clear$time_to_clear)
```


## Model of viral clearance

```{r}
cens_val = unique(CT_data$log10_cens_vl)
preds_mod1 = ifelse(colMeans(thetas_mod1$pred_log10_vl)<=cens_val,
                    cens_val,
                    colMeans(thetas_mod1$pred_log10_vl))
preds_mod2 = ifelse(colMeans(thetas_mod2$pred_log10_vl)<cens_val,
                    cens_val,colMeans(thetas_mod2$pred_log10_vl))
```



Plot data - patients with known vaccination status

```{r Fig3, fig.width=10}
cols = RColorBrewer::brewer.pal(n=5, 'Paired')[c(2,5)]
par(las=1, family='serif', cex.lab=1.3, cex.axis=1.3, mfrow=c(1,2))
out_surv=survfit(Surv(time_to_clear, clearance_event) ~ Trt_effect,
                 data = CT_data_time_clear)
plot(out_surv, conf.int=T, col=cols,lwd=2,
     xlab='Days since peak viral load',
     ylab='Proportion still positive')
grid()
mtext(text = 'a',side = 3,adj = 0,line = 1.5,cex = 1.3)
legend('bottomleft', col=cols,inset=0.03,
       lwd=3,legend = c('Not vaccinated','Vaccinated'))
out=survdiff(Surv(time_to_clear, clearance_event) ~ Trt_effect,
             data = CT_data_time_clear)
out$chisq

plot(CT_data$time, CT_data$log10_vl,
     ylim = c(0, 8), xlab='Days since peak viral load',
     ylab='RNA copies per mL', yaxt='n',
     panel.first=grid(), col=CT_data$Trt_effect+1)
axis(2, at = c(2,4,6,8), labels = c(expression(10^2),
                                      expression(10^4),
                                      expression(10^6),
                                      expression(10^8)))
mtext(text = 'b',side = 3,adj = 0,line = 1.5,cex = 1.3)
for(id in unique(CT_data$PersonID)){
  ind = CT_data$PersonID==id
  lines(CT_data$time[ind], 
        CT_data$log10_vl[ind], col=adjustcolor(cols[CT_data$Trt_effect[ind]+1],.6))
}

xx = aggregate(log10_vl~Day+Trt_effect, CT_data, median)
points(xx$Day[xx$Trt_effect==0], xx$log10_vl[xx$Trt_effect==0], pch=17, cex=1.5, col='darkblue')
points(xx$Day[xx$Trt_effect==1], xx$log10_vl[xx$Trt_effect==1],
       pch=17, cex=1.5,col='red')
legend('topright', title = 'Median daily value',
       legend = c('Not vaccinated', 'Vaccinated'),
       pch=17, col= c('darkblue','red'), inset=0.03)
```


## Spline fits

mgcv 

```{r gam_fits}
AUC_gam = array(dim = c(length(unique(CT_data$PersonID)), length(D_maxs)))
dat=data.frame(time = stan_data1$obs_day, 
               y=stan_data1$log_10_vl, 
               ID=as.factor(stan_data1$id))
mod_gam = gam(y ~ s(time) + s(ID, bs='re'), data = dat)
stan_data1$preds_gam = predict(mod_gam)

plot(stan_data1$obs_day, stan_data1$log_10_vl-stan_data1$preds_gam)
abline(h=0)

for(id in unique(stan_data1$id)){
  ind = which(stan_data1$id==id)
  xs = seq(0,14, length.out = 100)
  mypreds = predict(mod_gam,
                    data.frame(time=xs,ID=as.factor(id)))
  mypreds = mypreds/mypreds[1]
  ys = ifelse(mypreds<0, 0, mypreds)
  f = approxfun(x = xs, y = ys)
  for(ds in 1:length(D_maxs)){
    AUC_gam[as.numeric(id),ds] = 
      integrate(f = f, lower = D_min,
                upper = D_maxs[ds])$value
  }
}
```

## Estimated parameters


```{r parameter_estimates}
par(las=1, family='serif', cex.lab=1.5, cex.axis=1.5,mfrow=c(4,3))
cols = RColorBrewer::brewer.pal(n=3, 'Dark2')[1:2]

# A0
plot(density(thetas_mod1$intercept), main=expression('A'[0]),
     xlab='copies per mL (log10)', ylab='', yaxt='n', xlim = c(4.5, 6),
     panel.first=grid(),lwd=3, col=cols[1])
lines(density(thetas_mod2$intercept[,2]),lwd=3,lty=2, col=cols[2])

# B0 - model 2 only
plot(density(thetas_mod2$intercept[,1]), main=expression('B'[0]),
     xlab='copies per mL (log10)', ylab='', yaxt='n',
     panel.first=grid(),lwd=3,lty=2, xlim = c(0, 3), col=cols[2])

# alpha
plot(density(thetas_mod1$coef), main=expression(alpha), 
     xlab='Slope',
     ylab='', yaxt='n', xlim = c(0.5,1.2),
     panel.first=grid(),lwd=3, col=cols[1])
lines(density(thetas_mod2$coef[,2]),lwd=3,lty=2, col=cols[2])

# beta - model 2 only
plot(density(thetas_mod2$coef[,1]), main=expression(beta), 
     ylab='', yaxt='n', xlab='Slope',
     panel.first=grid(),lwd=3,lty=2, xlim = c(0, .4), col=cols[2])


# sigma 
plot(density(thetas_mod2$sigma_vl), main=expression(sigma),
     xlab='copies per mL (log10)', ylab='', 
     yaxt='n', xlim = c(0.6, 1),
     panel.first=grid(),lwd=3, lty=2, col=cols[2])
lines(density(thetas_mod1$sigma_vl),lwd=3, col=cols[1])

# degrees of freedom 
plot(density(thetas_mod2$t_dof), main='k',
     xlab='degrees of freedom', ylab='', 
     yaxt='n', xlim = c(4, 12),
     panel.first=grid(),lwd=3, lty=2, col=cols[2])
lines(density(thetas_mod1$t_dof),lwd=3, col=cols[1])


plot(density(thetas_mod1$sigmasq_u[,1]), main=expression(sigma[u1]),
     xlab='copies per mL (log10)', ylab='', yaxt='n', xlim = c(0.5, 1.2),
     panel.first=grid(),lwd=3, col=cols[1])
lines(density(thetas_mod2$sigmasq_u[,1]), lwd=3, lty=2, col=cols[2])

plot(density(thetas_mod1$sigmasq_u[,2]), main=expression(sigma[u2]),
     xlab='copies per mL (log10)', ylab='', yaxt='n', xlim = c(0.2, .7),
     panel.first=grid(),lwd=3, col=cols[1])
lines(density(thetas_mod2$sigmasq_u[,2]),lwd=3, lty=2, col=cols[2])

plot(density(thetas_mod2$sigmasq_u[,3]), main=expression(sigma[u3]),
     xlab='copies per mL (log10)', ylab='', 
     yaxt='n', xlim = c(0, 3),
     panel.first=grid(),lwd=3,lty=2, col=cols[2])

plot(density(thetas_mod2$sigmasq_u[,4]), main=expression(sigma[u4]),
     xlab='copies per mL (log10)', ylab='', yaxt='n', xlim = c(0, 1),
     panel.first=grid(),lwd=3,lty=2, col=cols[2])
```



Sensitivity analysis model 1

```{r sens_mod1}
thetas_mod1_sens = extract(out_mod1_sens)
par(las=1, family='serif', cex.lab=1.5, cex.axis=1.5,mfrow=c(2,3))
# A0
plot(density(thetas_mod1$intercept), main=expression('A'[0]),
     xlab='copies per mL (log10)', ylab='', yaxt='n', xlim = c(4.5, 5.5),
     panel.first=grid(),lwd=3, col=cols[1])
lines(density(thetas_mod1_sens$intercept))

# alpha
plot(density(thetas_mod1$coef), main=expression(alpha), 
     xlab='Slope',
     ylab='', yaxt='n', xlim = c(0.5,0.7),
     panel.first=grid(),lwd=3, col=cols[1])
lines(density(thetas_mod1_sens$coef))


# sigma 
plot(density(thetas_mod1$sigma_vl), main=expression(sigma),
     xlab='copies per mL (log10)', ylab='', 
     yaxt='n', xlim = c(0.8, 1),
     panel.first=grid(),lwd=3, col=cols[1])
lines(density(thetas_mod1_sens$sigma_vl))

# degrees of freedom 
plot(density(thetas_mod1$t_dof), main='k',
     xlab='degrees of freedom', ylab='', 
     yaxt='n', xlim = c(4, 12),
     panel.first=grid(),lwd=3,col=cols[1])
lines(density(thetas_mod1_sens$t_dof))


plot(density(thetas_mod1$sigmasq_u[,1]), main=expression(sigma[u1]),
     xlab='copies per mL (log10)', ylab='', yaxt='n', xlim = c(0.5, 1.2),
     panel.first=grid(),lwd=3, col=cols[1])
lines(density(thetas_mod1_sens$sigmasq_u[,1]))

plot(density(thetas_mod1$sigmasq_u[,2]), main=expression(sigma[u2]),
     xlab='copies per mL (log10)', ylab='', yaxt='n', xlim = c(0.2, .7),
     panel.first=grid(),lwd=3, col=cols[1])
lines(density(thetas_mod1_sens$sigmasq_u[,2]))
```


Sensitivity analysis model 2

```{r sens_mod2}
thetas_mod2_sens = extract(out_mod2_sens)

par(las=1, family='serif', cex.lab=1.5, cex.axis=1.5,mfrow=c(4,3))

# A0
plot(density(thetas_mod2$intercept[,2]), main=expression('A'[0]),
     xlab='copies per mL (log10)', ylab='', 
     yaxt='n', xlim = c(5, 6),lty=2,
     panel.first=grid(),lwd=3, col=cols[2])
lines(density(thetas_mod2_sens$intercept[,2]))

# B0 - model 2 only
plot(density(thetas_mod2$intercept[,1]), main=expression('B'[0]),
     xlab='copies per mL (log10)', ylab='', yaxt='n',
     panel.first=grid(),lwd=3,lty=2, xlim = c(0, 3), col=cols[2])
lines(density(thetas_mod2_sens$intercept[,1]))

# alpha
plot(density(thetas_mod2$coef[,2]), main=expression(alpha), 
     xlab='Slope',lty=2,
     ylab='', yaxt='n', xlim = c(0.8,1.3),
     panel.first=grid(),lwd=3, col=cols[2])
lines(density(thetas_mod2_sens$coef[,2]))

# beta - model 2 only
plot(density(thetas_mod2$coef[,1]), main=expression(beta), 
     ylab='', yaxt='n', xlab='Slope',
     panel.first=grid(),lwd=3,lty=2, xlim = c(0, .4), col=cols[2])
lines(density(thetas_mod2_sens$coef[,1]))


# sigma 
plot(density(thetas_mod2$sigma_vl), main=expression(sigma),
     xlab='copies per mL (log10)', ylab='', 
     yaxt='n', xlim = c(0.6, .82),
     panel.first=grid(),lwd=3, lty=2, col=cols[2])
lines(density(thetas_mod2_sens$sigma_vl))

# degrees of freedom 
plot(density(thetas_mod2$t_dof), main='k',
     xlab='degrees of freedom', ylab='', 
     yaxt='n', xlim = c(3, 9),
     panel.first=grid(),lwd=3, lty=2, col=cols[2])
lines(density(thetas_mod2_sens$t_dof))


plot(density(thetas_mod2$sigmasq_u[,1]), main=expression(sigma[u1]),
     xlab='copies per mL (log10)', ylab='', yaxt='n', xlim = c(0.5, 1.2),
     panel.first=grid(),lwd=3, col=cols[2],lty=2)
lines(density(thetas_mod2_sens$sigmasq_u[,1]))

plot(density(thetas_mod2$sigmasq_u[,2]), main=expression(sigma[u2]),
     xlab='copies per mL (log10)', ylab='',
     yaxt='n', xlim = c(0.2, .7),
     panel.first=grid(),lwd=3, lty=2, col=cols[2])
lines(density(thetas_mod2_sens$sigmasq_u[,2]))

plot(density(thetas_mod2$sigmasq_u[,3]), main=expression(sigma[u3]),
     xlab='copies per mL (log10)', ylab='', 
     yaxt='n', xlim = c(0, 3),
     panel.first=grid(),lwd=3,lty=2, col=cols[2])
lines(density(thetas_mod2_sens$sigmasq_u[,3]))

plot(density(thetas_mod2$sigmasq_u[,4]), main=expression(sigma[u4]),
     xlab='copies per mL (log10)', ylab='', yaxt='n', xlim = c(0, 1),
     panel.first=grid(),lwd=3,lty=2, col=cols[2])
lines(density(thetas_mod2_sens$sigmasq_u[,4]))

```



## Figure 1 - model fits

```{r Fig1}
mycols = RColorBrewer::brewer.pal(n=4, 'Dark2')
par(las=1, family='serif', cex.lab=1.5, cex.axis=1.5)
layout(mat = matrix(c(1,1,1,2,3,4),nrow = 2,byrow = T))
plot(CT_data$time, CT_data$log10_vl,yaxt='n',
     ylim = c(0, 8), xlab='Time since peak (days)',
     ylab='RNA copies per mL', panel.first=grid())
mtext(text = 'a',side = 3,adj = 0,line = 1.5,cex = 1.3)
for(id in unique(CT_data$PersonID)){
  ind = CT_data$PersonID==id
  lines(CT_data$time[ind], 
        CT_data$log10_vl[ind], col=adjustcolor('grey',.4))
}
axis(2, at = c(2,4,6,8), labels = c(expression(10^2),
                                    expression(10^4),
                                    expression(10^6),
                                    expression(10^8)))

xs = seq(0,14,length.out = 100)
preds1 = mean(thetas_mod1$intercept) - mean(thetas_mod1$coef)*xs
lines(xs, preds1, lwd=4, col=mycols[1])

preds2 = 
  apply(cbind(mean(thetas_mod2$intercept[,1])-
                mean(thetas_mod2$coef[,1])*xs,
              mean(thetas_mod2$intercept[,2])-
                mean(thetas_mod2$coef[,2])*xs),1,logSumExp)
lines(xs, preds2, lwd=4, col=mycols[2],lty=2)

preds3=predict(mod_gam, data.frame(time = xs, ID=-1), re.effect=NA)
lines(xs, preds3, lwd=4, col=mycols[3],lty=3)

xx = aggregate(log10_vl~Day, CT_data, median)
points(xx$Day, xx$log10_vl, pch=17, cex=2, col=mycols[4])

legend('topright',lwd=c(3,3,3,NA), 
       col=mycols,lty=c(1:3,NA),
       pch=c(NA,NA,NA,17), inset = 0.03, cex=1.3,
       legend = c('Exponential','Bi-exponential','Spline','Observed median'))

boxplot(stan_data1$log_10_vl-preds_mod1 ~ round(stan_data1$obs_day), 
        ylim = c(-2,2),xlab='Time since peak (days)',ylab='Residual error (log10 scale)',
        main='',panel.first=grid())
mtext(text = 'b: exponential',side = 3,adj = 0,line = 1.5,cex = 1.3)


boxplot(stan_data1$log_10_vl-preds_mod2 ~ round(stan_data1$obs_day),
        xlab='Time since peak (days)',ylab='Residual error (log10 scale)',
        main='',ylim = c(-2,2),panel.first=grid())
mtext(text = 'c: bi-exponential',side = 3,adj = 0,line = 1.5,cex = 1.3)

boxplot(stan_data1$log_10_vl-stan_data1$preds_gam~round(stan_data1$obs_day),
        xlab='Time since peak (days)',ylab='Residual error (log10 scale)',
        main='',ylim = c(-2,2),panel.first=grid())
mtext(text = 'd: spline regression',side = 3,adj = 0,line = 1.5,cex = 1.3)
```


Individual fits data 1 - we pick the individuals with the most data

```{r data1_individual_fits}
IDs_dense = as.numeric(names(which(sort(table(stan_data1$id))>=10)))
par(las=1, family='serif', cex.lab=1.3, cex.axis=1.3)
xs = seq(0,14, length.out = 100)
par(mfrow=c(4,4), mar=c(3,3,1,1),las=1)
for(id in IDs_dense){
  ind = which(stan_data1$id==id)
  plot(stan_data1$obs_day[ind], 
       stan_data1$log_10_vl[ind],
       ylim=c(0,8),
       xlab='', ylab='', main='', xaxt='n',yaxt='n',
       xlim=c(0,Max_Day), panel.first=grid())
  
  # Model 1
  intercept = median(thetas_mod1$intercept + thetas_mod1$theta_rand[,id,1])
  alpha = median(thetas_mod1$coef * exp(thetas_mod1$theta_rand[,id,2]))
  lines(xs, (intercept - alpha*xs), lty=1,lwd=2, col=mycols[1]) 
  
  # Model 2
  intercept1 = median(thetas_mod2$intercept[,2] + thetas_mod2$theta_rand[,id,1])
  intercept2 = median(thetas_mod2$intercept[,1] + thetas_mod2$theta_rand[,id,3])
  alpha = median(thetas_mod2$coef[,2]*exp(thetas_mod2$theta_rand[,id,2]))
  beta = median(thetas_mod2$coef[,1]*exp(thetas_mod2$theta_rand[,id,4]))
  ys = cbind(intercept1 - alpha*xs, intercept2 - beta*xs)
  preds = apply(ys, 1, logSumExp)
  lines(xs, preds,lwd=2,lty=2, col=mycols[2])
  
  # Spline model
  mypreds = predict(mod_gam,
                    data.frame(time=xs,ID=as.factor(id)))
  lines(xs, mypreds,lwd=2,lty=3, col=mycols[3])
  
  points(stan_data1$obs_day[ind],
         stan_data1$log_10_vl[ind],
         col=1, pch=16)
  axis(1, at = c(0,7,14))
  axis(2, at = c(2,4,6,8), labels = c(expression(10^2),
                                      expression(10^4),
                                      expression(10^6),
                                      expression(10^8)))
  
}
```




## Compute area under the curve

```{r auc, echo=F}
AUC_mod1 = AUC_mod2 = array(dim = c(stan_data1$n_id, length(D_maxs)))
FC_mod1 = FC_mod2 = array(dim = c(stan_data1$n_id, length(D_maxs)))
Nsim = 300

for(ds in 1:length(D_maxs)){
  writeLines(sprintf('Doing D max = %s ....', D_maxs[ds]))
  xs = seq(0,D_maxs[ds], length.out = 100)
  for(i in 1:stan_data1$n_id){
    # Model 0
    fcs = aucs = array(dim = Nsim); my_ix=1
    for(ss in sample(1:length(thetas_mod1$t_dof), Nsim, replace = F)){
      intercept = thetas_mod1$intercept[ss] + thetas_mod1$theta_rand[ss,i,1]
      alpha = thetas_mod1$coef[ss]*exp(thetas_mod1$theta_rand[ss,i,2])
      fcs[my_ix] = -alpha*D_maxs[ds]
      preds = 1 - alpha*xs/intercept # normalise by intercept
      f = approxfun(x = xs, y = ifelse(preds<0,0,preds))
      aucs[my_ix] = integrate(f = f, lower = D_min, upper = D_maxs[ds])$value
      my_ix=my_ix+1
    }
    AUC_mod1[i,ds] = median(aucs)
    FC_mod1[i,ds] = median(fcs)
    
    # Model 1
    fcs = aucs = array(dim = Nsim); my_ix=1
    for(ss in sample(1:length(thetas_mod2$t_dof), Nsim, replace = F)){
      
      intercept1 = thetas_mod2$intercept[ss,2] + thetas_mod2$theta_rand[ss,i,1]
      intercept2 = thetas_mod2$intercept[ss,1] + thetas_mod2$theta_rand[ss,i,3]
      alpha = thetas_mod2$coef[ss,2]*exp(thetas_mod2$theta_rand[ss,i,2])
      beta = thetas_mod2$coef[ss,1]*exp(thetas_mod2$theta_rand[ss,i,4])
      fcs[my_ix] = 
        logSumExp(c(intercept1, intercept2))-
        logSumExp(c(intercept1-alpha*D_maxs[ds],
                    intercept2-beta*D_maxs[ds]))
      ys = cbind(intercept1 - alpha*xs, intercept2 - beta*xs)
      preds = apply(ys, 1, logSumExp)
      preds_mod1 = preds/preds[1]
      f = approxfun(x = xs, y = ifelse(preds_mod1<0,0,preds_mod1))
      aucs[my_ix] = integrate(f = f, lower = D_min, D_maxs[ds])$value
      my_ix=my_ix+1
    }
    AUC_mod2[i,ds] = median(aucs)
    FC_mod2[i,ds] = median(fcs)
  }
}
```


Reduction in AUCs

```{r Fig2}
vacc_ind = unique(stan_data1$id[which(stan_data1$vacc==1)])
not_vacc_ind = unique(stan_data1$id[which(stan_data1$vacc==0)])

par(las=1, family='serif', cex.lab=1.3, cex.axis=1.3)
mycols = RColorBrewer::brewer.pal(n=3, 'Dark2')
plot(NA, NA, xlim = c(4, 15), ylim = c(-35,5),xaxt='n',
     xlab = 'Days follow-up for AUC', panel.first=grid(),
     ylab = 'Reduction in AUC: vaccinated versus not vaccinated')
axis(1, at = D_maxs)
abline(h=0, lty=2)
for(ds in 1:length(D_maxs)){
  out=(t.test(AUC_mod1[not_vacc_ind,ds], AUC_mod1[vacc_ind,ds]))
  points(D_maxs[ds]-.2, diff(out$estimate)/out$estimate[1]*100,pch=1,col=mycols[1],cex=1.5)
  lines(c(D_maxs[ds],D_maxs[ds])-.2,
        -out$conf.int/out$estimate[1]*100,col=mycols[1],lwd=3)
  writeLines(sprintf('Model 1: Wilcoxon test for AUC up to day %s: p= %s',D_maxs[ds], wilcox.test(AUC_mod1[not_vacc_ind,ds],AUC_mod1[vacc_ind,ds])$p.value))
  
  out=(t.test(AUC_mod2[not_vacc_ind,ds], AUC_mod2[vacc_ind,ds]))
  points(D_maxs[ds], diff(out$estimate)/out$estimate[1]*100,pch=2,col=mycols[2],cex=1.5)
  lines(c(D_maxs[ds],D_maxs[ds]),
        -out$conf.int/out$estimate[1]*100,col=mycols[2],lwd=3)
  writeLines(sprintf('Model 2: Wilcoxon test for AUC up to day %s: p= %s',
                     D_maxs[ds], 
                     wilcox.test(AUC_mod2[not_vacc_ind,ds],
                                 AUC_mod2[vacc_ind,ds])$p.value))
  
  out=(t.test(AUC_gam[not_vacc_ind,ds], AUC_gam[vacc_ind,ds]))
  points(D_maxs[ds]+.2, diff(out$estimate)/out$estimate[1]*100,pch=3,col=mycols[3],cex=1.5)
  lines(c(D_maxs[ds]+.2,D_maxs[ds]+.2),
        -out$conf.int/out$estimate[1]*100,col=mycols[3],lwd=3)
}
legend('bottomleft', inset=0.03, 
       legend = c('Exponential','Bi-exponential','Spline'),
       pch = 1:3, lwd=2, col=mycols)
```



```{r}
par(mfrow = c(2,2))
hist(colMeans(thetas_mod1$theta_rand[,not_vacc_ind,2]), breaks=seq(-1.5,1.5,by=.25))
hist(colMeans(thetas_mod1$theta_rand[,vacc_ind,2]), breaks=seq(-1.5,1.5,by=.25))

t.test(colMeans(thetas_mod1$theta_rand[,not_vacc_ind,2]),
       colMeans(thetas_mod1$theta_rand[,vacc_ind,2]))

```



## Variants and vaccines

effect of variant

```{r variant_effect, fig.width=8, fig.height=6}
par(las=1, family='serif', cex.lab=1.3, cex.axis=1.3)
par(mfrow=c(1,2),las=1)
ind_unique = !duplicated(stan_data1$id)
table(stan_data1$variant[ind_unique])

boxplot(apply(thetas_mod1$theta_rand[,,2],2,mean) ~ stan_data1$variant[ind_unique],
        xlab='',las=2,
        ylab = 'Slope difference',varwidth=T)


boxplot(apply(thetas_mod2$theta_rand[,,2],2,mean) ~ stan_data1$variant[ind_unique],
        xlab='',las=2,
        ylab = 'Slope difference',varwidth=T)
```


effect of vaccine

```{r vaccine_effect, fig.width=8, fig.height=6}
par(las=1, family='serif', cex.lab=1.3, cex.axis=1.3)
par(mfrow=c(1,2),las=1)
table(stan_data1$vacc[ind_unique])

boxplot(apply(thetas_mod1$theta_rand[,,2],2,mean) ~ stan_data1$vacc[ind_unique],xlab='Vaccinated',
        ylab = 'Slope difference',varwidth=T)

boxplot(apply(thetas_mod2$theta_rand[,,2],2,mean) ~ stan_data1$vacc[ind_unique],
        xlab='Vaccinated',ylab = 'Slope difference',varwidth=T)

plot(stan_data1$obs_day, stan_data1$log_10_vl,
     xlim=c(0,7), ylim=c(0,8), type='n',main='Exponential',
     xlab='Days since peak viral load',ylab='CT',
     panel.first=grid())
for(i in 1:stan_data1$n_id){
  # Model 0
  intercept = mean(thetas_mod1$intercept + thetas_mod1$theta_rand[,i,1])
  alpha = mean(thetas_mod1$coef + exp(thetas_mod1$theta_rand[,i,2]))
  preds = 1 - alpha*xs/intercept
  vacc = stan_data1$vacc[ind_unique][i]
  if(!is.na(vacc)){
    if(vacc==0) {mycol = 'grey';mylty=2} else {mycol = 'red';mylty=2}
    lines(xs, preds*intercept, col = mycol,lty=mylty,lwd=2)
  }
}
legend('topright',col=c('grey','red'),inset = 0.03,lty=c(2,1),
       legend = c('Not vaccinated','Vaccinated'),lwd=2)

plot(stan_data1$obs_day, stan_data1$log_10_vl,
     xlim=c(0,7), ylim=c(0,8), type='n',main='Bi-exponential',
     xlab='Days since peak viral load',ylab='CT',
     panel.first=grid())
for(i in 1:stan_data1$n_id){
  # Model 1
  intercept = mean(thetas_mod2$intercept + thetas_mod2$theta_rand[,i,1])
  intercept = mean(thetas_mod2$intercept + thetas_mod2$theta_rand[,i,3])
  alpha = mean(thetas_mod2$coef[,1] + exp(thetas_mod2$theta_rand[,i,2]))
  beta = mean(thetas_mod2$coef[,2] + exp(thetas_mod2$theta_rand[,i,4]))
  ys = cbind(intercept - alpha*xs, intercept - beta*xs)
  preds = apply(ys, 1, logSumExp)
  vacc = stan_data1$vacc[ind_unique][i]
  if(!is.na(vacc)){
    
    if(vacc==0) {mycol = 'grey';mylty=2} else {mycol = 'red';mylty=1}
    lines(xs, preds, col = mycol,lty=mylty,lwd=2)
  }
}

```


## Sample size estimation

```{r Fig4, fig.height=7, fig.width=8}
res = read.csv('Rout/all_sims_RateClearance.csv')
res2 = read.csv('Rout//all_sims_TimeToClear.csv')
res2$power = as.numeric(res2$pval< 0.025)
thetas_mod2 = extract(out_mod2)
set.seed(75873426)

# simulate data to illustrate underlying simulation process
sim_vl1 =
      sim_individuals(thetas = thetas_mod2,
                      t_design = seq(0,7,length.out = 20),
                      Trt_effect = c(rep(1,1000),rep(1.5,1000)),
                      p_before_peak = 1,
                      LOD = 1)
sim_vl2 =
      sim_individuals(thetas = thetas_mod2,
                      t_design = seq(0,7,length.out = 20),
                      Trt_effect = c(rep(1,1000),rep(1.5,1000)),
                      p_before_peak = 0,
                      LOD = 1)
xx_1 = aggregate(True_Log_VL ~ time+Trt_effect, sim_vl1, mean)
xx_2 = aggregate(True_Log_VL ~ time+Trt_effect, sim_vl2, mean)
ind1 = xx_1$Trt_effect==1
ind2 = xx_2$Trt_effect==1


par(las=1, family='serif', cex.lab=1.3, cex.axis=1.3)
layout(mat = matrix(c(1,2,1,3),nrow = 2,byrow = T))

col_peak = RColorBrewer::brewer.pal(n = 8,name = 'Set2')[c(8,6)]
plot(xx_1$time[ind1], xx_1$True_Log_VL[ind1], type='l',
     xlab='Time since enrolment (days)', ylab='RNA copies per mL',
     panel.first=grid(), yaxt='n',lwd=2,col=col_peak[1])
lines(xx_1$time[!ind1], xx_1$True_Log_VL[!ind1],lty=2,lwd=2,col=col_peak[1])

lines(xx_2$time[!ind2], xx_2$True_Log_VL[!ind2],lty=2,lwd=2,col=col_peak[2])
lines(xx_2$time[ind2], xx_2$True_Log_VL[ind2],lwd=2,col=col_peak[2])
legend('topright',col=c(col_peak,1,1),lwd=2,lty = c(1,1,1,2),legend = c('Before peak','After peak','No drug','Drug (50% increase)'),inset = 0.03)
axis(2, at = c(2,3,4,5), labels = c(expression(10^2),
                                    expression(10^3),
                                    expression(10^4),
                                    expression(10^5)))

mtext(text = 'a', side = 3,line = 1.5, adj = 0, cex = 1.3)
t_names = c('Twice daily (7d)',
            'Twice daily (5d)',
            'Once daily (7d)',
            'Once daily (5d)')
my_titles = c('b - 30% increase','c - 50% increase')
my_lty = c(1,2,1,2)
cols = RColorBrewer::brewer.pal(n = 8,
                                name = 'Dark2')[c(1,2,4,8)]

for(tt in c(1.3,1.5)){
  xx=aggregate(power ~ N + t_design,
               data = res[res$effect==tt, ], mean)
  plot(xx$N, 100*xx$power, ylim = c(0,100),
       panel.first=grid(),col = cols[xx$t_design],pch=16,
       xlab= ' Sample size (per arm)', ylab = 'Power (%)')
  for(j in unique(xx$t_design)){
    ind = xx$t_design==j
    lines(xx$N[ind], 100*xx$power[ind], 
          col = cols[j],lwd=2)
  }
  
  xx2=aggregate(power ~ N + t_design,
               data = res2[res2$effect==tt, ], mean)
  for(j in unique(xx2$t_design)){
    ind = xx2$t_design==j
    lines(xx2$N[ind], 100*xx2$power[ind], 
          col = cols[j],lwd=2, lty=2)
  }
  mtext(text = my_titles[which(tt==unique(res$effect))], 
        side = 3,line = 1.5, adj = 0, cex = 1.3)
  if(tt==1.3){
    
legend('topleft', legend = c('Rate of clearance','Time to clearance'),
       lty=1:2,col = 1, inset=0.03,lwd=2)
  }
}
legend('bottomright', legend = t_names,
       col = cols, inset=0.03,lwd=2)


aggregate(power ~ effect, data = res, mean)
aggregate(power ~ t_design, data = res[res$effect>1,], mean)

range(aggregate(power ~ t_design+effect+N,res,length)$power)
table(aggregate(power ~ t_design+effect+N,res2,length)$power)
```


